#Investigation of Distributions  
### Author: Jeffrey Thatcher  
[Directions][1]

[1]:https://class.coursera.org/statinference-035/human_grading/view/courses/975168/assessments/4/submissions  

##Overview  
The goal of this report is to use the central limit theorem to show that an 
estimation of the mean of a probability density function, specifically the 
exponential density function (EDF) in this case, is a good approximation of the 
calculated theoretical mean of the density function.  

In the following experiment, we will estimate the mean of the EDF by taking 1000
random samples of the EDF in 40 different experiments. The average of these 40
different experiments will be calculated and it is expected that this value is 
very close to the theoretical mean of the EDF.

lambda is 0.2 for all the following experiments and the exponential density
function has the following appearance:

```{r warning=F, fig.width=3.5, fig.height=2.5, echo=FALSE}
library(ggplot2)
library(stats)
set.seed(777)
lambda = 0.2

#Plot of the exponential density function with line at the mean
x = seq(0,15,0.01)
p1 = qplot(x, dexp(x,lambda), geom="line", ylab="f(x)") + 
        geom_vline(xintercept = 5, color="green")
p1
```

##Simulations  
The sample mean will be simulated by taking 1000 random samples of the 
exponential distribution in 40 separate experiments. From each experiment we 
will calculate the sample mean (represents the population mean) and then average
these 40 sample means to estimate the theoretical mean.

To begin we generate a matrix with 40 columns that each contain 1000 random
samples of the EDF.

```{r}
DF <- as.data.frame(matrix(1,1000,40))
DFexp <- sapply(DF,rexp, rate=0.2)
```

Next we calculate the mean and variance for each column and store these for the
next sections.

```{r}
#Mean of each experiment
DFmean <- as.data.frame(cbind(seq(1,40,1),colMeans(DFexp)))
colnames(DFmean) <- c("Experiment","Mean")

#Variance of each experiemnt
DFvar <- apply(DFexp,2,var)
DFvar <- as.data.frame(cbind(seq(1,40,1), DFvar))
colnames(DFvar) <- c("Experiment","Variance")
```

##Sample Mean versus Theoretical Mean  
Sample Mean versus Theoretical Mean: Include figures with titles. In the figures, highlight the means you are comparing. Include text that explains the figures and what is shown on them, and provides appropriate numbers.

The EDF function is f(x) = lambda{e}^{-lambda*x} for x >= 0. Lambda is the rate 
parameter. The theoretical mean of the EDF distribution is 1/lambda and in 
this case is 1/0.2 or `r 1/lambda`.

From our 40 experiments where we randomly sampled the EDF 1000 times, we calculated
the mean.  These 40 means are represented as a histogram in the following plot:

```{r fig.width=3.5, fig.height=2.5, echo=FALSE}
ggplot(DFmean, aes(x=Mean)) + 
        geom_histogram(fill = "skyblue", color = "black", size = 0.2, binwidth=0.1)
```

**If we calculate the average of these means we obtain the value: `r colMeans(DFmean[2])`**  

The value `r colMeans(DFmean[2])` is very close to the theoretical mean
`r 1/lambda`. This is not surprising and shows that this experiment did agree
with the central limit theorem.

##Sample Variance versus Theoretical Variance  
We expect the same tendency of the experimentally calculated mean to be true for
the experimentally calculated variance of the EDF.  The theoretical Variance of 
the EDF can be calculated as the square of the standard deviation, and the standard 
deviation of the EDF is calculated as 1/lambda.  Therefore, the variance for the
EDF is 5^2 or `r (1/lambda)^2`.

From our 40 experiments where we randomly sampled the EDF 1000 times, we calculated
the variance for these experiments individually and present them in the
following histogram:

```{r fig.width=3.5, fig.height=2.5, echo=FALSE}
ggplot(DFvar, aes(x=Variance)) + 
        geom_histogram(fill = "darkorchid2", color = "black", size = 0.2, binwidth=1)
```

**If we calculate the average of these variances we obtain the value: `r colMeans(DFvar[2])`**  

The value `r colMeans(DFvar[2])` is very close to the theoretical variance
`r (1/lambda)^2`. This is not surprising and shows that this experiment did agree
with the central limit theorem.

##Distribution  
One can tell the distribution of Means gathered from our 40 experiments  is
approximately normal by an informal approach of plotting a normal distribution 
over the histogram.  In this case the y-axis must be normalized to 1 prior to 
mapping the normal distribution to this figure.  From the following figure, 
we can tell that the histogram approximates a normal distribution.

```{r fig.width=3.5, fig.height=2.5, echo=FALSE}
stdev = sd(DFmean[,2])
ggplot(DFmean, aes(x=Mean)) + 
        geom_histogram(aes(y=..density..),fill = "skyblue", color = "black", size = 0.2, binwidth=0.1) +
        stat_function(fun = dnorm, args = list(mean = 5, sd = stdev))
```

The same principle could be applied to the distribution of variances.
