---
title: "Analysis of ToothGrowth data"
author: "J. Thatcher"
output: word_document
---
Load the ToothGrowth data
```{r warning=FALSE, echo=FALSE}
library(ggplot2)
library(xtable)
options(xtable.floating = FALSE)
source("multiplot.R")

TGdata <- ToothGrowth
```

There are three variables in the toothgrowth dataset. They are length (len, levels = continuous),
supplement (supp, levels = OJ and VC), and dose (dose, levels = 0.5, 1.0, and 2.0). We can quickly look to see any correlation between the variables by pairwise plots of each variable.

```{r warning=F, echo=F, fig.width=6.5, fig.height=2.3}
p1 <- ggplot(TGdata, aes(supp,len)) + geom_point(size = 3, color="green")
p2 <- ggplot(TGdata, aes(dose,len)) + geom_point(size = 3, color="skyblue") + 
        geom_smooth(method = "lm")
p3 <- ggplot(TGdata, aes(supp,dose)) + geom_point(size = 3,color="red")

multiplot(p1,p2,p3,cols=3)
```

There appears to be positive correlation between the dose and length variables. 
Which means that length of the tooth can be predicted by the dose. Furthermore,
the supp variable also seems to have a minor correlation with len. This becomes
more obvious when we separate the dose ~ len plot by the factor supp in the
following plot:

```{r warning=F, echo=F, fig.width=5, fig.height=2.5}
p4 <-ggplot(TGdata, aes(x= dose,y= len, colour=supp)) + geom_point(size = 3) + 
        geom_smooth(method = "lm")
p4
```

##Hypotheses
The following hypotheses are relevant to this dataset:

For the supp variable  

H0: the effect of supp OJ = the effect of supp VC on tooth length  
H1: the effect of supp OJ != the effect of supp VC on tooth length

For all pairwise comparisons of the dose variable  

H0: the effect of dose A = the effect of dose B on tooth length  
H1: the effect of dose A != the effect of dose B on tooth length

For all pairwise comparisons of the supp variable within a single level of the
dose variable  

H0: the effect of supp OJ = the effect of supp VC on tooth length  
H1: the effect of supp OJ != the effect of supp VC on tooth length  

##The effect of supp
For the first set of hypotheses:  
H0: the effect of supp OJ = the effect of supp VC on tooth length  
H1: the effect of supp OJ != the effect of supp VC on tooth length  

First we will ask, does the mean of the supp OJ fall within the mean of the supp
VC?  

We can quickly calculate the 95% confidence interval with the `t.test` function,
as well as the p-value for two-tailed comparison.  
```{r warning=F, echo=F, }
t.test(TGdata[1:30,1], TGdata[31:60,1], var.equal = TRUE)
```
We can interpret this results table in such a way that the 95% confidence
interval for the difference in means does include zero and the p-value is slightly
larger than 0.05. This suggests that we cannot reject H0 and rule-out
the possibility that the difference in means was a result of chance.  

##The effect of dose
For all pairwise comparisons of the dose variable  

There are 3 total pariwise comparisons.  
0.5 vs. 1.0  
0.5 vs. 2.0  
1.0 vs. 2.0  

Results of these t-test can be quickly calculated, but we will consider that the
possibility of getting a significant comparison increases with the number of tests
performed on the data and therefore we need to apply a correction to alpha. We will
use the Bonferroni correction, which is conservative and simple to implement.  

```{r warning=F, echo=F}
d1 <- c(TGdata[1:10,1], TGdata[31:40,1])
d2 <- c(TGdata[11:20,1], TGdata[41:50,1])
d3 <- c(TGdata[21:30,1], TGdata[51:60,1])
AvB <- t.test(d1, d2, var.equal = TRUE)
AvC <- t.test(d1, d3, var.equal = TRUE)
BvC <- t.test(d2, d3, var.equal = TRUE)

results <- as.data.frame(rbind(AvB$p.value, AvC$p.value, BvC$p.value))
results <- cbind(c("0.5 vs. 1.0 ","0.5 vs. 2.0 ","1.0 vs. 2.0 "),results)
colnames(results) <- c("| Comparison of Dose |", "|  P-Value   |")
results
```
In this case we find that the p values for all three comparisons are extremely
small, and even with the conservative Bonferroni correction where alpha =`r 0.05/3`
we can still reject the null hypotheses and agree that the increase the variable
len increases with increasing levels of the variable dose, regardless of the
level of the variable supp.  

##supp vs len at each level of dose
finally, we will consider the effect of either supplement when we remove the
effect of the dose. To do this, we will compare the two levels of supp, OJ and VC
at each of the three levels of dose. Again, the Bonferroni correction will be
used.  

There are 3 total pariwise comparisons.  
Dose 0.5:  VC vs. OJ  
Dose 1.0:  VC vs. OJ  
Dose 2.0:  VC vs. OJ  

possible comparisons and their results are as follows:  

```{r warning=F, echo=F}
VCd1 <- c(TGdata[1:10,1])
VCd2 <- c(TGdata[11:20,1])
VCd3 <- c(TGdata[21:30,1])

OJd1 <- c(TGdata[31:40,1])
OJd2 <- c(TGdata[41:50,1])
OJd3 <- c(TGdata[51:60,1])

A1vB1 <- t.test(VCd1, OJd1, var.equal = TRUE)
A2vB2 <- t.test(VCd2, OJd2, var.equal = TRUE)
A3vB3 <- t.test(VCd3, OJd3, var.equal = TRUE)

results2 <- as.data.frame(rbind(A1vB1$p.value, A2vB2$p.value, A3vB3$p.value))
results2 <- cbind(c("Dose 0.5:  VC vs. OJ","Dose 1.0:  VC vs. OJ",
                              "Dose 2.0:  VC vs. OJ"),round(results2,5))
colnames(results2) <- c("|Comparison of Supp by Dose |", "| P-Value |")
results2
```
Here we find that the p-values for the first two levels of dose are less than
the alpha value we selected (0.05/3 or `r 0.05/3`). The third comparison, however,
is not less than alpha = `r 0.05/3`. Therefore, we conclude that the response of
len to supp is dependent on the level of dose and that the supp OJ is associated
with higher values of len at the doses 0.5 and 1.0.  

##Assumptions
Our overall assumptions for these tests include:  

1. groups of data were normally distributed  
2. equal variances within each of the t-tests  

#Appendix: R-code  
```{r eval=FALSE}
library(ggplot2)
library(xtable)
options(xtable.floating = FALSE)
source("multiplot.R")

TGdata <- ToothGrowth

p1 <- ggplot(TGdata, aes(supp,len)) + geom_point(size = 3, color="green")
p2 <- ggplot(TGdata, aes(dose,len)) + geom_point(size = 3, color="skyblue") + 
        geom_smooth(method = "lm")
p3 <- ggplot(TGdata, aes(supp,dose)) + geom_point(size = 3,color="red")

multiplot(p1,p2,p3,cols=3)

p4 <-ggplot(TGdata, aes(x= dose,y= len, colour=supp)) + geom_point(size = 3) + 
        geom_smooth(method = "lm")
p4

t.test(TGdata[1:30,1], TGdata[31:60,1], var.equal = TRUE)

d1 <- c(TGdata[1:10,1], TGdata[31:40,1])
d2 <- c(TGdata[11:20,1], TGdata[41:50,1])
d3 <- c(TGdata[21:30,1], TGdata[51:60,1])
AvB <- t.test(d1, d2, var.equal = TRUE)
AvC <- t.test(d1, d3, var.equal = TRUE)
BvC <- t.test(d2, d3, var.equal = TRUE)

results <- as.data.frame(rbind(AvB$p.value, AvC$p.value, BvC$p.value))
results <- cbind(c("0.5 vs. 1.0 ","0.5 vs. 2.0 ","1.0 vs. 2.0 "),results)
colnames(results) <- c("| Comparison of Dose |", "|  P-Value   |")
results

VCd1 <- c(TGdata[1:10,1])
VCd2 <- c(TGdata[11:20,1])
VCd3 <- c(TGdata[21:30,1])

OJd1 <- c(TGdata[31:40,1])
OJd2 <- c(TGdata[41:50,1])
OJd3 <- c(TGdata[51:60,1])

A1vB1 <- t.test(VCd1, OJd1, var.equal = TRUE)
A2vB2 <- t.test(VCd2, OJd2, var.equal = TRUE)
A3vB3 <- t.test(VCd3, OJd3, var.equal = TRUE)

results2 <- as.data.frame(rbind(A1vB1$p.value, A2vB2$p.value, A3vB3$p.value))
results2 <- cbind(c("Dose 0.5:  VC vs. OJ","Dose 1.0:  VC vs. OJ",
                              "Dose 2.0:  VC vs. OJ"),round(results2,5))
colnames(results2) <- c("|Comparison of Supp by Dose |", "| P-Value |")
results2
```